Catastrophic Goodhart: regularizing RLHF with KL divergence does not mitigate heavy-tailed reward misspecification

### 2024-10-24 paper/main.tex
add Drake affiliation (with placeholder email)
address reparameterizing reward

% ready for submission
\usepackage{neurips_2024}

@@

\author{%
  Thomas Kwa
  Independent / FAR Labs
  \texttt{kwathomas0@gmail.com} \\
  \And
  Adrià Garriga-Alonso \\
  FAR AI \\
  \texttt{adria@far.ai} \\
}

@@

Recall that KL divergence between two distributions P and Q is defined as \[
D_{\mathrm{KL}}(P \| Q)=\sum_{x \in \mathcal{X}} P(x) \log \left(\frac{P(x)}{Q(x)}\right)
\].

@@

\section{Experimental Methodology}

Our theoretical results now raise the question of whether the error in reward models is heavy-tailed or light-tailed in practice. \footnote{Note that distributions over a finite set are bounded and cannot be heavy-tailed in a technical sense, and models with a finite context window have a finite input space. We say that a distribution of reward or error is heavy-tailed if it is well-modeled by a heavy-tailed distribution on its support.} If we observe the reward distribution to be light-tailed, this is a strong indication that error is light-tailed. \footnote{It is possible for $U$ to be light-tailed while $X$ and $V$ are both heavy-tailed, but this is unusual and we do not expect it to happen in practice.}

@@

\section{Results\label{sec:experiments}}

@@

@@

\citet{gao2023scaling} found that optimizing the reward of small reward models causes overoptimization: a decrease in utility with increasing optimization. However, we observed that reward models are light-tailed, and (Theorem \ref{thm4}) that independence combined with light-tailed error prevents overoptimization. We think this discrepancy is explained by dependence between error and utility. Policies optimized for high error may activate features in the proxy reward models that are undesirable according to the true utility function.\footnote{There are other explanations possible. Perhaps better optimization methods would find heavy-tailed reward in open reward models; or OpenAI's reward models have heavy-tailed error (and their results are straightforwardly explained by our Theorem~\ref{thm1}), while open reward models have light-tailed error.}
More research is needed to understand why high-error completions have low utility and to design reward models that do not suffer from this problem; Perhaps it is possible to construct reward models whose errors are in directions orthogonal to human preferences, so that the large-reward completions do not have lower utility.

>>>>

% ready for submission
\usepackage[final]{neurips_2024}

@@

\author{%
  Thomas Kwa \\
  Independent / FAR Labs \\
  \texttt{kwathomas0@gmail.com} \\
  \And
  Drake Thomas \\
  Anthropic \\
  \texttt{TODO} \\
  \And
  Adrià Garriga-Alonso \\
  FAR AI \\
  \texttt{adria@far.ai} \\
}

@@

Recall that KL divergence between two distributions P and Q is defined as \[
D_{\mathrm{KL}}(P \| Q)=\sum_{x \in \mathcal{X}} P(x) \log \left(\frac{P(x)}{Q(x)}\right).
\]

@@

\section{Experiments}

Our theoretical results now raise the question of whether the error in reward models is heavy-tailed or light-tailed in practice. \footnote{Note that distributions over a finite set are bounded and cannot be heavy-tailed in a technical sense, and models with a finite context window have a finite input space. For our purposes, a distribution of reward or error is effectively heavy-tailed if it takes on sufficiently large values and is well-modeled by a heavy-tailed distribution on its support.} If we observe the reward distribution to be light-tailed, this is a strong indication that error is light-tailed. \footnote{It is possible for $U$ to be light-tailed while $X$ and $V$ are both heavy-tailed, but this is unusual and we do not expect it to happen in practice.}

@@

\subsection{Results\label{sec:experiments}}

@@

\subsection{Reparameterizing reward}

In some cases, a heavy-tailed reward can be reparameterized to make it light-tailed and avoid catastrophic Goodhart; however, in settings where the true reward is heavy-tailed, making reward artificially light-tailed or bounded can result in unintended behavior.

For example, a stock-trading agent should be rewarded by profit, but financial returns are known to be heavy-tailed. If we clip or otherwise transform rewards into a bounded interval, it will have no incentive to take into account huge gains or losses. Since RLHF rewards as implemented in Ziegler et al are unbounded, clipping or transforming rewards could itself cause reward misspecification.

In some cases, e.g. when the reward is not the true intended one, it is possible to reparameterize the reward without adverse effects. In the RL literature for Atari games, rewards are changes in score clipped to $[-1, 1]$ \citep{machado2018revisiting}.

@@

\citet{gao2023scaling} found that optimizing the reward of small reward models causes overoptimization: a decrease in utility with increasing optimization. However, we observed that reward models are light-tailed, and (Theorem \ref{thm4}) that independence combined with light-tailed error prevents overoptimization. We think this discrepancy is explained by dependence between error and utility. Policies optimized for high error may activate features in the proxy reward models that are undesirable according to the true utility function.\footnote{There are other explanations possible. Perhaps better optimization methods would find heavy-tailed reward in open reward models; or OpenAI's reward models have heavy-tailed error (and their results are straightforwardly explained by our Theorem~\ref{thm1}), while open reward models have light-tailed error.} % TODO: discuss this with PPO experiment
More research is needed to understand why high-error completions have low utility and to design reward models that do not suffer from this problem; perhaps it is possible to construct reward models whose errors are in directions orthogonal to human preferences, so that the large-reward completions do not have lower utility.
